---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Langston Alexander"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup, message=FALSE}
#1
library(tidyverse)
library(corrplot)
library(RColorBrewer)
library(lubridate)
library(agricolae)

LakeData <- read.csv(
  "C:/Users/lwa8/Documents/R/ENV872/Environmental_Data_Analytics_2022/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",
                     stringsAsFactors = TRUE)
as.Date(LakeData$sampledate, format = "%m/%d/%y")

#2
mytheme <- theme_gray(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature in July does not change with depth in at least one lake.
Ha: Mean lake temperature in July changes with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, warning=FALSE, message=FALSE}
#4
processedLakeData <- 
  LakeData %>%
  filter(month(sampledate) == 07)%>%
  select(c(lakename, year4, daynum, depth, temperature_C))%>%
  na.omit()

#5
ggplot(processedLakeData, aes(x = depth, y = temperature_C))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm")+
  ylim(0, 35)+
  labs(title = "Temperature by Depth in July", x = "Depth (M)", y = "Temperature(C)")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It looks as though depth has an impact on temperature in a lake until, starting at about 5 meters deep, the effects on temperature level off. This trend could be slightly biased because the majority of data points are taken at shallow depths and realtively few are taken past 10 meters. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
TempByDepth <- lm(data = processedLakeData, temperature_C ~ depth)
summary(TempByDepth)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: A one meter decrease in depth is associated with an approximately 1.9 degrees Celsius temperature decrease. This result is statistically significant with a p-value < 0.01, an adjusted R-squared of 0.7387, and 27500 on 1 and 9726 degrees of freedom. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

LakeTempAIC <- lm(data = processedLakeData, temperature_C ~ year4 + daynum + depth)
summary(LakeTempAIC)
step(LakeTempAIC)

#10

MLRLakeTemp <- lm(data = processedLakeData, temperature_C ~ year4 + daynum + depth)
summary(MLRLakeTemp)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests that year4, daynum, and depth all be included in the multiple regression. This improves the observed variance (R-squared) by about 0.02. This means that by adding year and day that the sample was taken, we are coving .2% more of the observed variance in temperature that by just looking at depth. This is a slight improvement over only using depth as the explanatory variable but not by much. Depth is still the overwhelmingly the most dominant factor in predicting water temperature. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
LakeTempAnova <- aov(data = processedLakeData, temperature_C ~ lakename)
summary(LakeTempAnova)
LakeTempAnovaLM <- lm(data = processedLakeData, temperature_C ~ lakename)
summary(LakeTempAnovaLM)

plot(LakeTempAnova)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Because we have a very low p-value < 0.01, we can reject the null hypothesis that all the lakes have the same average temperature in July and accept the alternative hypothesis that at least one lake has a different average temperature in July. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

ggplot(processedLakeData, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.5)+
  labs(y = "Temperature (C)", x = "Depth (M)", 
       color = "Lake Name", 
       title = "Temperature by Depth in July")+
  ylim(0,35)+
  scale_color_brewer(palette = "Set1")+
  geom_smooth(aes(color = lakename,), method = "lm", se = F)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(LakeTempAnova)

TempMeanTotals <- HSD.test(LakeTempAnova, "lakename", group = TRUE)
TempMeanTotals

TempMeanTotalsPlot <- ggplot(processedLakeData, aes(x = lakename, y = temperature_C)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("a", "ab", "e", "de", "c", "c", 
                         "de", "bc", "d")) +
  labs(x = "Lake Name", y = "Temperature (C)", title = "Temperature Quartiles by Lake") +
  ylim(0, 35)
print(TempMeanTotalsPlot)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically speaking, Paul and Ward Lakes have the same mean temperature as Peter Lake. No lake's average temperature is completely distinct from all the others. For every lake there is at least one other lake with a statistically similar mean temperature. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could also use a two-sample t-test to investigate whether Peter and Paul Lakes have distinct mean temperatures. 


