---
title: "Assignment 09: Data Scraping"
author: "Langston Alexander"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(rvest)
library(lubridate)
library(purrr)

mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"))
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2020 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2

webpage <- 
  read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
pswid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
month <- c(1,5,9,2,6,10,3,7,11,4,8,12)
year <- 2020
maxuse.df <- data.frame(
  "Month" = as.factor(month),
  "Max_Withdrawl" = as.numeric(max.withdrawals.mgd),
  "System_Name" = water.system.name,
  "Ownership" = ownership,
  "PSWID" = pswid,
   'Date' = my(paste(month,"-",year)))


maxuse.df <- arrange(maxuse.df, Month)

maxuse.df <- maxuse.df %>%
  mutate(
    MonthName = c(
      "Jan", "Fed", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"
  ))

#5

ggplot(maxuse.df, aes(x = Month, y = Max_Withdrawl, group = 1))+
  geom_line(size = 1.2)+
  geom_point(color = "red")+
  labs(title = "Max Monthly Water Withdrawls from Durham in 2020", 
       y = "Max Water Withdrawl (MGD)")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.

maxuse.scrape <- function(PSWID, Year){
  
  #Fetch Website
  
  webpage <- read_html(paste0(
    'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
    PSWID, '&year=', 
    Year))
  
  #Scrape element data
  
  water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
pswid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()


  #Convert to df

  month <- c(1,5,9,2,6,10,3,7,11,4,8,12)
  maxuse <- data.frame(
  "Month" = as.factor(month),
  "Max_Withdrawl" = as.numeric(max.withdrawals.mgd),
  "System_Name" = water.system.name,
  "Ownership" = ownership,
  "PSWID" = pswid,
   'Date' = my(paste(month,"-",Year)))


  maxuse <- arrange(maxuse, Month)
  
  return(maxuse)
  
}

maxuse.scrape('03-32-010', 2019)

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r}
#7

maxuseDurham.df <- maxuse.scrape("03-32-010", 2015)

ggplot(maxuseDurham.df, aes(x = Month, y=Max_Withdrawl, group = 1))+
   geom_line(size = 1.2)+  
  geom_point(color = "red")+
    labs(title = "Max Monthly Water Withdrawls from Durham in 2015", 
         y = "Max Water Withdrawl (MGD)")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Durham.2015.data}
#8
maxuseAshville.df <- maxuse.scrape("01-11-010", 2015)
maxuseAshville.df

DurAsheMaxUSe.df <- bind_rows(maxuseDurham.df, maxuseAshville.df)

ggplot(DurAsheMaxUSe.df, aes(x = Month, y = Max_Withdrawl, fill = System_Name))+
  geom_bar(position = 'dodge', stat = "identity")+
  labs(
    title = "Max Water Withdrawl of Durham and Asheville by Month in 2015", 
    y = "Max Water Withdrawl (MGD)",
    fill = "City Name")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9

the_years <- rep(2010:2019)
the_pswid <- "01-11-010"

ashw_dfs <- map(the_years, maxuse.scrape, PSWID = the_pswid)

ashe_df <- bind_rows(ashw_dfs)

ashe_df$year <- as.factor(year(ydm(ashe_df$Date)))

ggplot(ashe_df, aes(x = Date, y = Max_Withdrawl))+
  geom_point()+
  geom_smooth(color = "red", se = F)+
  labs(title = "Max Monthly Water Withdrawls from Asheville 2010-2019", 
       y = "Max Water Withdrawl (MGD)")

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

It looks like there wasn't a clear trend in water usage from 2010 until 2017. In 2017 water usage started to rise and continued to do so through 2019. 